% THIS IS AN EXAMPLE DOCUMENT FOR VLDB 2012
% based on ACM SIGPROC-SP.TEX VERSION 2.7
% Modified by  Gerald Weber <gerald@cs.auckland.ac.nz>
% Removed the requirement to include *bbl file in here. (AhmetSacan, Sep2012)
% Fixed the equation on page 3 to prevent line overflow. (AhmetSacan, Sep2012)

\documentclass{vldb}
\let\proof\relax
\let\endproof\relax

\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, draftwatermarkbottom, draftwatermarktop, amsthm}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
\usepackage{xspace}

\newcommand{\etal}{{\em et al.}\xspace}
\newcommand{\etc}{{\em etc}\xspace}
\newcommand{\cf}{{\em c.f.,}\xspace}
\newcommand{\versus}{{\em vs.}\xspace}
\newcommand{\ie}{{\em i.e.,}\xspace}
\newcommand{\eg}{{\em e.g.,}\xspace}

\theoremstyle{definition}
\newtheorem{observation}{Observation}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}


\renewcommand{\ttdefault}{fi4}

\usetikzlibrary{positioning, snakes, shapes}

\newenvironment{myitemize}
{

   \vspace{0mm}
    \begin{list}{$\bullet$}{\leftmargin=1em}
        \setlength{\topsep}{0em}
        \setlength{\parskip}{0pt}
        \setlength{\partopsep}{0pt}
        \setlength{\parsep}{0pt}         
        \setlength{\itemsep}{.25em} 
        \setlength{\itemindent}{0em}
}
{
    \end{list} 
    \vspace{-.5em}
}

\newenvironment{myenumerate}
{

   \vspace{0mm}
   \newcounter{qcounter}
    \begin{list}{\arabic{qcounter}.~}{\usecounter{qcounter}\leftmargin=1em}
        \setlength{\topsep}{0em}
        \setlength{\parskip}{0pt}
        \setlength{\partopsep}{0pt}
        \setlength{\parsep}{0pt}         
        \setlength{\itemsep}{.25em} 
        \setlength{\itemindent}{0em}
}
{
    \end{list} 
    \vspace{-.5em}
}

\begin{document}


% ****************** TITLE ****************************************

\title{Highly Available Transactions: Virtues and Limitations}
{\author{Peter Bailis{\fontsize{12}{14}$^\dagger$}, Aaron Davidson{\fontsize{12}{14}$^\dagger$}, Alan Fekete{\fontsize{12}{14}$^\diamond$}, Ali Ghodsi{\fontsize{12}{14}$^{\dagger,\ddagger}$}, Joseph M. Hellerstein{\fontsize{12}{14}$^\dagger$}, Ion Stoica{\fontsize{12}{14}$^\dagger$}\\{{\fontsize{12}{14}$^\dagger$}\hspace{.5mm}UC Berkeley\hspace{4mm}{\fontsize{12}{14}$^\diamond$}\hspace{.5mm}University of Sydney\hspace{4mm}{\fontsize{12}{14}$^\ddagger$}\hspace{.5mm}KTH/Royal Institute of Technology}}}
\maketitle

\begin{abstract}
To minimize network latency and remain online during partial failure
and network partitions, many modern distributed data storage systems
eschew transactional functionality, which provides strong semantic
guarantees for groups of multiple operations over multiple data
items. In this work, we consider the problem of providing Highly
Available Transactions (HATs), or transactional guarantees that do not
suffer from unavailability during system partitions or incur high
network latency.  We introduce a new taxonomy of high availability and
analyze a existing ACID and distributed consistency guarantees to
identify which can and cannot be achieved in HAT systems. In doing so,
we analytically and experimentally quantify the benefit of providing
high availability as well as the associated semantic compromises.
\end{abstract}

In the debate over NoSQL systems versus traditional DBMS technologies for distributed data management, one key issue is that many NoSQL systems aim for availability in the face of partitions, and provide no, or restricted, support for transactions that group multiple operations over multiple data items. The authors recently advocated Highly Available Transaction (HAT) systems, where general transactions are provided with ACID-like semantic guarantees such as weak isolation levels, in ways that remain available during partitions and avoid high network latency. Here we introduce a new taxonomy of partition-availability properties. We analyse existing ACID and distributed consistency guarantees, showing the availability that can be achieved for each.  We analytically and experimentally estimate the performance gains and semantic compromises of providing high availability, compared to traditional serializable transactions in an internet-scale setting.

\input{intro.tex}

\input{motivation.tex}

\input{model.tex}

\input{hat-analysis.tex}

\input{evaluation.tex}

\input{relatedwork.tex}

\input{futurework.tex}

\input{conclusion.tex}


% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
% vldb_sample.bib is the name of the Bibliography in this case
\footnotesize
\bibliography{hat-vldb}  
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references

%APPENDIX is optional.
% ****************** APPENDIX **************************************
% Example of an appendix; typically would start on a new page
\pagebreak

\begin{appendix}
\input{availability-appendix.tex}

\input{hat-appendix.tex}
\end{appendix}
\end{document}
